\ttitle

### 2013-05-08 Chapters/Chapter1.tex
Create outlines for new focus: temporal segmentation

% Chapter 1

\chapter{Introduction} % Main chapter title

@@

\lhead{Chapter 1. \emph{Introduction}} % This is for the header on each page -
%perhaps a shortened title

@@

\section{Problem statement}
*** Give context problems, means to reach (e.g. monitoring of patients,  getting conclusions over health and activity).
%----------------------------------------------------------------------------------------

\input{./Chapters/Chapter1/literature_review}

\section{System architecture}

*** [GRAPHICS diagram: sensors --> signal pre-processing --> feature extraction --> analysis [* segmentation, * labeling] --> conclusions]

>>>>

% !TEX root = ../main.tex
% Chapter 1

\chapter{Introduction}

@@

\lhead{Chapter 1. \emph{Introduction}} % This is for the header on each page - perhaps a shortened title

@@

\section{Outline}
\begin{itemize}
  \item Context of research (human activity recognition), real-world applications
  \item Current methods, wrapper vs. filter methods
  \item Problem statement with current filter methods (which follows from Chapter \ref{Chapter3} which goes in-depth with methods).
  \item Purpose of this research. E.g. "Find a better algorithm for short-activity segmentation"
  \item Relate to real-world applications
\end{itemize}

### 2013-05-08 Chapters/Chapter2.tex
Extended bib with sensor papers
Create outlines for new focus: temporal segmentation

% Chapter 2
% !TEX root = ../main.tex

\chapter{Techniques} % Main chapter title

\label{Chapter2} % For referencing the chapter elsewhere, use \ref{Chapter1}

\lhead{Chapter 2. \emph{Techniques}} % This is for the header on each
%page -
%perhaps a shortened title

%----------------------------------------------------------------------------------------

\section{Signal pro-processing and sensor fusion for timed patterns}

*** [fusion of accelerometer data, gyroscopes, magnetic field sensors,
pressure, microphones, etc]

\section{Temporal Segmentation}

This section will give an introduction and in-depth analysis of temporal
segmentation.

\subsection{Aims of segmentation}
When processing and analyzing time series of data, e.g. motion measurements,
stock market fluctuations or natural language, first a low-level division
between the discriminative parts of the stream must be made. One can view this
as splitting the series into the \emph{atoms}, which are the building blocks
of the total stream. These building blocks will be the aggregation of
non-overlapping, internally homogeneous segments \cite{himberg2001time}. This
means that the data points inside a segment should have some resemblance
relation to each other and their difference lies between some boundary. The
process of segmenting can be viewed as a subproblem to context analysis of
time series. Temporal segmentation is closely related to temporal clustering,
although it is a stricter, and simpler, process. Whereby clustering only
restricts the data points on their distance relation (as used in a Voronoi
diagram), within a segment the data points must also be contiguous.

The task of segmentation can be performed in a manual
matter, by cutting and labeling parts of the stream into coherent parts. This
would require human (expert) knowledge and does not yield a clear cut because
of ambiguity. With increasing storage abilities and easier motion capture
systems, there is a desire for automated systems which perform the
segmentation task unsupervised. Some algorithms used have the (often desired)
side-effect of also clustering the segments, such that classes of segments can
be discovered in the time series. These algorithms would not only be able to
make a distinction between walking, sitting and walking, but would also
recognize the reappearance of the walking activity.

\subsection{Formal definition}
Formally, temporal segmentation is dividing a time series $s$, which consists
of $N$ samples $\mathbf{x}(1),\mathbf{x}(2),\dots,\mathbf{x}(N)$ from
$\mathbf{R}^d$. Individual \emph{segments} are referenced by $s(a,b)$,
consisting of the consecutive samples
$\mathbf{x}(a),\mathbf{x}(a+1),\dots,\mathbf{x}(b)$, $a \le b$. Let $s_1 =
s(a,b)$ and $s_2 = s(b+1,c)$ be two segments, then their concatenation is
$s_1s_2 = s(a,c)$. A segmentation $S$ of $s$ consists of a sequence of $k$
non-empty segments $s_1s_2 \dots s_k = s$.
This notation is adopted from \cite{himberg2001time}.

As stated, informally each segment should be internally homogeneous. This can
formally be measured with an cost function $F$, indication the heterogeneity
of a segment. The overall aim is to minimize the cost $F$. The cost of a
segment is a function from the data points and the number of data points $n =
b - a + 1$ and is expressed as
\begin{equation} \label{eq:segment_cost}
\mathrm{cost}_F (s(a,b)) = F(\mathbf{x};n|\mathbf{x} \in s(a,b))
\end{equation}
The cost of a \emph{k-segmentation} $S$ is the summation of the costs of the
$k$ segments:
\begin{equation} \label{eq:segmentation_cost}
\mathrm{Cost}_F (s_1 s_2 \dots s_k) = \sum_{i=1}^{k} \mathrm{cost}_F (s_k)
\end{equation}
With the objective of minimizing the cost function, the optimal
$k$-segmentation $S_F^\mathit{opt}(s;k)$ is the segmentation with minimal
$\mathrm{Cost}_F(s_1 s_2 \dots s_k)$ over all possible $k$-segmentations.

The cost function, to calculate the heterogeneity of a (set of) segment(s),
can
be any function. A simple and natural function would be the sum of variances
of the segments. The overall cost function would then be
\begin{equation} \label{eq:cost_variances}
\mathrm{Cost}_V = \frac{1}{N} \sum_{i=i}^{k} \sum_{j=c_{i-1}+1}^{c_i} \|
\mathbf{x}(j) - \mu_i \|^2
\end{equation}
where $\mu_i$ is the mean vector of data points in segment $s_i$.

\subsection{Application in research fields}

[CHARACTERISTICS OF HUMAN MOTION]
temporal variability, invariance over time, metrics over actions.

[COMPUTER VISION]

[GRAPHICS/VIDEO]

[DATA-MINING]

[MODEL BASED]

To analyze time series it is often preferred to divide the stream in segments
of correlated data. After dividing, each segment represent a period in time in
which the same activity is performed. Or, stated otherwise, it results in
transitions moments between activities.

\subsection{PCA Based Methods}

Many fields of research have been active in the unsupervised segmentation of
data. Many authors rely on a form of Principal Component Analysis (PCA), as
used a.o. in \cite{barbivc2004segmenting}. Often PCA is used to reduce the
dimensionality of the data being processed [REFERENCE] by only using the top
$r$ dimensions to describe the data set. It is observed that data series of
simple motions have a lower dimensionality then complexer motions. When a
simple (repetitive) motion is about to end and fluently transforms in a new
motion, there will be a window of time in which a high dimensionality will be
present, due to the new motion. After this period of transition, the
dimensionality will decrease, since only the new simple motion is present in
the window of time. The first algorithm of \cite{barbivc2004segmenting} is
based on this principle.

Given a set of data points, a lower dimensional hyperplane can by constructed
to which the data points can be projected. This projection on a lower
dimension introduces a error to the original position. When the error is
fixed, less dimensions are needed for simple motions in which movements of
body parts are highly correlated. For segments in which the data points are
lesser correlated, e.g. because of transition state, a higher degree of
dimensions of the hyperplane is needed to represent the data with equal error
degree. When the dimensionality is reduced from $d$ to $r \le d$, the ratio of
error
$E_r$ can be calculated as

\begin{equation}
	E_r = \frac{\sum_{j=1}^{r} \sigma_j^2}{\sum_{j=1}^{d} \sigma_j^2}
\end{equation}

where $\sigma_j$ are the singular values as a result from Singular Value
Decomposition (SVD), which is closely related to PCA \cite{shlens2005tutorial}.

In \cite{barbivc2004segmenting} the stream of frames is analyzed on cut-frames
to find transitions between action. First, for a number of $k$ frames (e.g.
the
equivalent of 2.5 seconds) the required dimensionality $r$ to keep the error
$E_r$
below some threshold $\tau$ is calculated. This will yield in an error $e_i$
for the first
$i$ frames. When more frames are added to the window, the error will increase
with a low constant when it is still in the same activity, due to noise in the
activity. When a new activity
starts, e.g. at frame $j$, the error at frame $j$ will start increasing
faster. This can be expressed by the derivative of the error rate $d_i = e_i -
e_{i-l}$, where $l$ is a constant to remove noise. From this derivative the
mean and standard deviation can be calculated, for each point. When a
derivative $d_j$ rises more than a factor $k_\sigma = 3$ standard deviations
from the mean, a transition point is encountered. The previous frames are then
cut from the sequence (as a segment) and the algorithm starts over.

*** [figure to illustrate derivative and standard deviation]

A second approach in \cite{barbivc2004segmenting} uses the probabilistic
variant of PCA (PPCA) to model the data set as a Gaussian distribution instead
of ignoring the frames which do not fit in the subspace. Over windows of
frames the mean and variance are calculated. In a forward manner the
Mahalanobis distance of a new window of frames is calculated, which represents
the likelihood of the new window belonging to the same segment as the original
widow. When the distance decreases, the likelihood increases which happens
when the motions in the becomes more homogeneous. When a peak in the distance
is reached, the new window of frames indicates a heterogeneous collection of
motions in the window and thus a low likelihood of membership and a indication
of a transition. In order to distinct activities and subactivities (which
require a subset of motions is a distinct activity) the algorithm is also
processed backward over the data series.

*** [figure to illustrate mahalanobis distance measure and peaks]

The third algorithm in \cite{barbivc2004segmenting} is based on the
observation that data points (frames) tend to form clusters in the space.
These clusters are represented by $k$ Gaussian distributions for which each
the Expectation-Maximization (EM) algorithm estimates the mean $m_j$,
covariance matrix $\sum_{j}$ and prior $\pi_j$. With all the Gaussian
distributions estimated, the data points are assigned to the cluster with the
highest membership likelihood. When two consecutive frames $x_i$ and $x_{i+1}$
belong to different clusters, a transition of activities is recognized. Note
that this algorithm succeeds in segmenting the data and also labels the
similar simple activities.

A drawback in this system, and many others which implement a variant of the
$k$-means algorithm, is that the number of clusters $k$ need to be
predetermined. To cope with this, often the algorithm is performed multiple
times for different values of $k$. Using some criteria, e.g. the Bayesian
Information Criterion \cite{pelleg2000x} or the Davies-Bouldin Index which
guides $k$-means clustering as used in \cite{krause2003unsupervised}.

\subsection{Statistical methods}
An method to process data, or create a basis for further processing, is by
analyzing the statistical properties of a data set. When there is a continuous
stream of data, it is possible to extract events by comparing the statistical
properties of sliding time windows. By definition, events have a different
statistical profile then their background. When two consecutive actions are
regarded as the background, then a transition is an event between them.

In \cite{guenterberg2009automatic} a automatic system is used to segment a
continuous stream of data. By calculating features as activity level from the
standard deviation of a sliding window and applying Adaptive Thresholds, the
data can be segmented in active and rest parts. There is no model
representation.

\subsection{Hidden Markov Models Based Methods}

\subsection{Bayesian Methods}

\subsection{Principal Component Analysis}
When working with simple 2- or 3-dimensional data it is often, for humans,
easy to discover patterns in the set. The data can be plotted and the lines or
planes among which the data points lie gives an indication of the pattern.
With Principal Component Analysis (PCA) this process is also possible for
automated systems. With this method the overall form of a set of data points
can be represented, the most discriminative dimensions can be found, the
dimensionality of a set can be reduced (which yield in data compression) or
the result can be used to characterize and differentiate sets of data points.

The following steps are performed, which are explained below and illustrated
in figure *** [add figure with plot of sets]:

\begin{enumerate}
	\item \textbf{Gather data} and represent them in a chosen number of
	features,
	\item \textbf{Subtract mean} to center the data around the axis. The new
	data set will have mean zero,
	\item \textbf{Generate covariance matrix} by calculating all pairwise
	feature variances,
	\item \textbf{Calculate Eigenvectors and Eigenvalues} of the covariance
	matrix. The Eigenvectors are then normalized to make further calculations
	easier,
	\item \textbf{Generate feature vector} which indicated which features are
	characteristic or meant to keep in de data set, by comparing the
	Eigenvalues.
\end{enumerate}

The workings of the PCA \cite{smith2002tutorial} relies on the concepts of
standard deviation, variance,
covariance, eigenvectors and eigenvalues of matrices. These concepts will be
discussed very briefly \footnote{For a more in-depth discussion we would like
to refer the reader to \cite{jolliffe2005principal}}. The standard deviation
and variance
of a set are measures for the
spread around the mean for a single dimension, or feature. The covariance
between two features (dimensions of the data points) indicates how they are
related; when positive the two
features will increase together; when negative one will decrease when the
other increases and when zero they are unrelated. The covariance matrix gives
all the covariances for all pairs of features. This matrix is symmetrical
about the diagonal and the values on the diagonal are the variances for each
feature. For this matrix the eigenvectors can be computed. Eigenvectors have
the characteristic that when they multiply a matrix, the resulting vector is a
multiple of the Eigenvector. The amount by which it is the multiple is the
Eigenvalue. This is illustrated in formulae \ref{eq:no-eigenvector} and
\ref{eq:eigenvector}. The second formula shows an Eigenvector
$ \left( \begin{smallmatrix}
3 \\ 2 \end{smallmatrix} \right)$ and its
associated Eigenvalue, 4. The vector in the first formula is not an
Eigenvector.

\begin{equation} \label{eq:no-eigenvector}
	\begin{pmatrix} 2 & 3 \\ 2 & 1 \end{pmatrix}
\times
	\begin{pmatrix} 1 \\ 3 \end{pmatrix}
=
	\begin{pmatrix} 2 \cdot 1 + 3 \cdot 3 \\ 2 \cdot 1 + 1 \cdot 3
	\end{pmatrix}
=
	\begin{pmatrix} 11 \\ 5 \end{pmatrix}
\end{equation}

\begin{equation} \label{eq:eigenvector}
	\begin{pmatrix} 2 & 3 \\ 2 & 1 \end{pmatrix}
\times
	\begin{pmatrix} 3 \\ 2 \end{pmatrix}
=
	\begin{pmatrix} 2 \cdot 3 + 3 \cdot 2 \\ 2 \cdot 3 + 1 \cdot 2
	\end{pmatrix}
=
	\begin{pmatrix} 12 \\ 8 \end{pmatrix}
=
	4 \times \begin{pmatrix} 3 \\ 2 \end{pmatrix}
\end{equation}

It are these Eigenvectors and Eigenvalues that makes PCA possible. Given a
matrix of size $n \times n$ (only square matrices have Eigenvectors), $n$
Eigenvectors can be found. Each Eigenvector is perpendicular, or orthogonal,
to the others. The Eigenvectors combined describes the lines over which the
data is plotted. The Eigenvector with the largest Eigenvalue is the
\emph{principal} component of the data set. It is said that it is the most
significant feature. Thus the Eigenvectors, and thereby the features or
dimensions of the data set, can be sorted on significance by the Eigenvalues.

With this ordering there are a few applications possible. The first is to just
make a (comprehensible) representation of the data. The Eigenvectors describe
the cloud of data points and thus are a compressed representation of the data.
When the original data points are to be compressed, it is possible to remove
the least significant features and reconstruct the data points from the
resulting Eigenvectors. This will yield in a lossy compression. *** [figure to
illustrate]. An extension of this application is to determine the number of
features needed to keep the compression within a certain error criterion, as
used in \cite{barbivc2004segmenting}. Sets of data can then be distinguished
by the number of features needed to describe the points.

*** [graph, a bit of formulae with sets and dimensions]

%-------------------

\subsection{Hidden Markov Models}
Many dynamic systems can be viewed as some sequence of consecutive states
which are observable by the world. These kind of systems are known as Markov
Chains. When the state in which the system is can not be observed with
complete certainty, a system known as a Hidden Markov Model (HMM) can be
constructed. It has been shown that HMMs are effective on in the task of
speech recognition \cite{rabiner1989tutorial}. Because of some resemblance
between natural speech construction and human activities, modifications have
been able to classify human activities \cite{guenterberg2009distributed} and
extensions are implemented which create layers of HMMs to handle different
granularity over time and reduce retraining
\cite{oliver2002layered,perdikis2008recognition}. A HMM is a state-space model
representing a single model for each label which needs to be classified.

Formally, a Hidden Markov Model is characterized by the following elements
\cite{rabiner1989tutorial}:
\begin{enumerate}
	\item A number of $N$ states in the model. These states are a
	representation of the underlying mechanism of the system and can not be
	directly observed from outside (hence the \emph{hidden} model). Individual
	states are referenced by $S = \{S_1, S_2, \dots, S_N \}$ and a single
	state in time $t$ is denoted by $q_t$.
	\item An alphabet of $M$ discrete observable symbols for each state. An
	observed symbol is the only indication of the systems current situation.
	Each individual symbol is referenced by $V = \{ v_1, v_2, \dots, v_m \}$.
	\item A state transition probability distribution $A = \{ a_{ij} \}$
	indication the probability of being in state $j$ at time $t+1$ when the
	state on time $t$ was $i$:
	\begin{eqnarray}
		a_{ij} = P [ q_{t+1} = S_j | q_t = S_i ], & 1 \le i, j \le N
	\end{eqnarray}
	Note that by setting the probability $a_{ij}=0$ for some $i$ and $j$
	indicates that $j$ is not reachable from state $i$.
	\item A symbol observation probability distribution in state $j$, $B = \{
	b_j(k) \}$, where
	\begin{eqnarray}
	b_j(k) = P[ v_k\ \mathrm{at} \ t | q_t = S_j], & 1 \le j \le N \nonumber \\
	& 1 \le k \le M
	\end{eqnarray}
	Note that every symbol can be observed in every state.
	\item The initial state distribution in which the system starts $\pi= \{
	\pi_i \}$ where
	\begin{eqnarray}
	\pi_i = P[q_1 = S_i], & 1 \le i \le N
	\end{eqnarray}
\end{enumerate}
This system is often abbreviated to
\begin{equation}
	\lambda = \{ \pi_i, a_{ij}, b_j(k)\}
\end{equation}
The HMM will act as a generator giving a sequence of $T$ observations, each
being a symbol from $V$:
\begin{equation}
	O = O_1, O_2 \dots O_T
\end{equation}

Given the characteristics of a specific HMM, there are three major application
which serve the classification of an activity. The first application is, given
a sequence of observations $O = O_1,O_2,\dots,O_T$, to label this sequence
with the activity with
the highest associated probability for that label, e.g. compute $P(O|\lambda)$
for each model $\lambda$. The second task a HMM can perform is to
\emph{explain} a observation sequence $O$, by choosing an optimal
corresponding state sequence $Q = q_1,q_2,\dots,q_T$. The last application is
more of a problem statement, being the task to train the HMM to adjust the
parameters $\lambda = \{ \pi_i, a_{ij}, b_j(k)\}$ to maximize $P(O|\lambda)$.
In the overall task of activity recognition, one would first use the last
application to train a set of HMMs with the provided and labeled training
data. The second application can be used to further optimize the HMM (set the
number of states, adjust vector codebook etc.) and improve the capability of
modeling the activities. Finally, when the system is provided with unlabeled
data the first application is used to identify the observed sequence with
trained examples, by comparing the probability result of multiple HMMs.

Generally a HMM can be fully connected, or ergodic model. Other configurations
are also possible, of which the left-right model, or Bakis model, is popular
for temporal applications \cite{rabiner1989tutorial}. This because the
structure of the state probability distribution resembles the construction of
possible states over time. This left-right model can be generalized by setting
\begin{eqnarray}
	a_{ij} = 0, & j > i + \Delta
\end{eqnarray}
for some $\Delta$, indicating the possible \emph{distance} of the jumps over
states.

*** [INSERT GRAPHICS SHOWING STRUCTURE OF HMM]

*** [discrete and continuous HHMs]

*** Applications of HHM: \cite{shi2009towards} uses 10 coefficients extracted
with FFT from 6 axis (3 times accelerometer and 3 times gyroscope). Example of
continuous HMM (?). Recognition rates fo 50\%-100\%.

In \cite{oliver2002layered} it is stated that HMMs are robust mechanisms with
respect to changes in temporal segmentation of observations, although they
suffer from a few problems when applied to reason about longer and complexer
sequences over time. They state that, especially with limited training data,
the HMM tend to be overfitted and tend to lack structure and have an excess of
parameters. The overcome this limitation, they introduce a Layered Hidden
Markov Model (LHHM) which they show is more robust to temporal changes in
observations. They make use of the hypothesis from the field of psychology,
that human behavior is hierarchically structured \cite{zacks2001event}. In
\cite{perdikis2008recognition} a layered construction of HMMs is used which
feeds the low-level data from a sliding window to the lower layer. This layer
will recognize simple motions, such as moving the arm in a pre-defined
segment. The results from these lower HMMs are then fed to a higher-level
layer of HMMs which are able to recognize abstract motions, such as adjusting
a monitor and picking up a pen. The results showed that the layered system was
better at recognition then a similar single-layered setup. It was also robust
in the sense that when the environment was changed, only the relatively simple
low-level HMMs needed to be retrained.

*** [DIAGRAM WITH EXAMPLE OF LOW- AND HIGH-LEVEL HMMs, SUCH AS MOVING ARM,
PICKING UP PEN]

*** Viterbi algorithm / path; find the most likely sequence of cause-states
for
the observed measures.

%-------------------

\subsection{Segmentation as clustering}
In the previous section the discussed methods all relied on the stream of data
points and tried to find cuts the discriminate between successive different
type of activities. An other approach is to consider the tasks of segmentation
as a type of clustering \cite{zhou2008aligned}. In clustering the objective is
to assign labels, or classes, to all the data points indication a similar type
of activity. A clustering $\mathcal{L}$ is thereby more informative then a
segmentation but is also harder to produce.

A clustering $\mathcal{L}$ is generated from a sequence of elements
$\mathbf{X}$ which is decomposed in $m$ disjoint segments, each belonging to
one of the $k$ classes. A segment $\mathbf{Y}_i \hat{=}
\mathbf{X}_{[s_i,s_{i+1})}$ is composed of frames from position $s_i$ to
$s_{i+1}$. A vector $g_{ci} = 1$ indicates class membership if $\mathbf{Y}_i$
belongs to class $c$, otherwise $g_{ci} = 0$.

When regarding segmentation of human motion as a task of clustering the
difficulty is to model the temporal variability of actions
and defining a robust metric between temporal actions. To overcome this,
\cite{zhou2008aligned} introduces Aligned Cluster Analysis (ACA), by minimizing
\begin{equation} \label{eq:ACA}
J_{\mathit{ACA}}(\mathbf{G},\mathbf{s}) =
\sum_{c=i}^{K} \sum_{i=1}^{m} g_{ci} \mathit{dist}_c
(\mathbf{X}_{[s_i,s_{i+1})})
\end{equation}

The characteristic of ACA is that is enables segments to span over different
number of data points, whereas the standard kernel $k$-means algorithm results
in equally sized segments. [!!! TRUE?!] The second difference is that the
kernel used in $\mathit{dist}_c$ to measure the distance from a segment to the
class which it is assigned to uses the Dynamic Time Alignment Kernel
[REFERENCE?] to measure between time series.

%----------------------------------------------------------------------------------------

\section{Clustering}

This section will give an introduction and overview to clustering of data.
When processing time series of data the line of distinction between
segmentation and clustering is very fine. This section will introduce
clustering for the purpose as used in this project.

Data clustering can be used with two different applications, exploratory or
confirmatory \cite{jain1999data}. In the former, the purpose is to discover
clusters in a point cloud of data points. Here the cluster is defined as a
group of homogeneous data points measured by some coherence measure, often a
distance function with the data points being defined as vectors. The produced
clusters, which together form a representation of the data examined, can be
used in the
latter application to assign new provided data points to a cluster and thereby
classify them.

Depended on the precise setup, clustering can be used as a successive step
after or an implementation of temporal segmentation. When temporal
segmentation is used to create successive coherent data points, clustering can
be used to recognize the same activities in different points of time. The data
points provided to the clustering will be some representation of each segment
and each segment will be labeled with a activity. Another mechanism could be
to extract features from the raw data points and use these to create clusters
directly *** [temporal modifications]. Both mechanisms requires the
exploratory and confirmatory phases.

The exploratory phase roughly consists of the following five steps
\cite{jain1999data}
\begin{enumerate}
	\item data point representation,
	\item definition of coherence,
	\item grouping data points,
	\item cluster abstraction,
	\item output qualification
\end{enumerate}

The first step is to find a \emph{representation} of the data points and
clusters, considering the number, type and scale of the features and the
number of desired classes to cluster in. When dealing with high-dimensional
data points a \emph{feature selection} can be performed to use only the most
discriminating features. When the raw features are not useful enough,
\emph{extraction} can be used to create new synthetic features. ***
[qualitative versus quantitative/conceptual]

When the data points are represented in a meaningful way, the measure of
\emph{coherence} between pairs of points must be defined. Often this is
implemented as the Euclidean
or Mahalanobis distance when the data points are represented as vectors or
other similarity measures for conceptual patterns.

In the most critical step, the \emph{grouping}, the data points with the
highest coherence are linked together to form a cluster. The result can be a
hard partition over the data or a fuzzy membership degree to each cluster for
each data point. When applying rules for merging and splitting cluster, a
hierarchical partition is constructed. Partitional clustering algorithms
assign data points to clusters by optimizing some criterion, e.g. the mean
squared distance from each data point to the clusters' centroid.

Especially in case of large data sets, the resulting clusters will be defined
by many data points. To form a compacter and simpler representation,
\emph{abstraction} can be used. This will simply analysis by humans or
automated systems. A common abstraction is to use the clusters' centroid
\cite{diday1976clustering} or parameters of Gaussian patterns.

The final step which can be applied to the resulting partition is some
\emph{qualification}. The quality of multiple partitions can be compared and
the validity of the partition can be determined. A partition is valid if
reasonably it could not been constructed by change or as some random process.
External references of models can be used to compare, or the internal data
points can be examined. *** [better wordings]

%-------------------

\subsection{Formal definition}

%-------------------

\subsection{Application is research fields}

%----------------------------------------------------------------------------------------

\section{Temporal pattern recognition}

\subsection{Dynamic Time Warping}
Used to measure similarity between time sequences. Exact matching is
high-cost, so approximations such as Minimum Bounding Rectangles are used.

\subsection{k-means clustering}
[EXPLAIN] divide data set $n$ into $k$ clusters.

Among many unsupervised clustering techniques, $k$-means is successfully
applied to large data sets. It is simple to implement and linear in time
complexity so computationally attractive \cite{jain1999data}. A drawback of
the method is that the results of the algorithm greatly depends on the initial
configuration (the data points which will act as centroids) and the number of
cluster $k$ must be determined beforehand.

Generally, the $k$-means methods will minimize the squared error for a
clustering $\mathcal{L}$ criterion which is defined as the distance from the
data points the centroid for each cluster in $\mathcal{K}$. This is expressed
as optimizing to a local optimum the energy function
\begin{equation} \label{eq:k-means energy}
e^2(\mathcal{K},\mathcal{L}) =
\sum_{j=i}^{K}\sum_{i=1}^{n_j}\|\mathbf{x}_i^{(j)} -
\mathbf{c}_j\|^2
\end{equation}

There are several limitation on the $k$-means method. One of these is that
only spherical shapes of cluster can be generated. One of the extensions is
kernel $k$-means \cite{scholkopf1998nonlinear}, which implicitly projects the
data points to a higher dimension and thereby is able to form irregular shaped
cluster.

\subsection{Self-organizing Map}

\subsection{Support Vector Machine}

\subsection{Na\"{i}ve Bayes}

\section{Unsupervised clustering of temporal patterns}

>>>>

% !TEX root = ../main.tex
% Chapter 2

\chapter{Literature review}

\label{Chapter2} % For referencing the chapter elsewhere, use \ref{Chapter2}

\lhead{Chapter 2. \emph{Literature review}} % This is for the header on each page - perhaps a shortened title

%----------------------------------------------------------------------------------------
\section{Outline}
\begin{itemize}
  \item Literature review about Temporal Segmentation (previous draft was more about classification)
  \item Consider methods for the context of filter-methods for classification
  \item Take a loot at 3-4 different kind of methods for change detection:
    \begin{itemize}
      \item Dimensionality reduction
      \item Density-ratio estimation
      \item Support Vector Machines (?) - if there are more sources about this
      \item CUSUM - or other more traditional methods
    \end{itemize}
  \item With each method, shortly look at characteristics, strengths and weaknesses and consider applicability to accelerometer sensor data
\end{itemize}

### 2013-05-08 Chapters/Chapter3.tex
Create outlines for new focus: temporal segmentation

% Chapter 3

\chapter{Experiments} % Main chapter title

\label{Chapter3} % For referencing the chapter elsewhere, use \ref{Chapter1}

\lhead{Chapter 3. \emph{Experiments}} % This is for the header on each 
%page - 
%perhaps a shortened title

%----------------------------------------------------------------------------------------

\section{}

>>>>

% !TEX root = ../main.tex
% Chapter 3

\chapter{Change detection by Density-Estimation}

\label{Chapter3} % For referencing the chapter elsewhere, use \ref{Chapter3}

\lhead{Chapter 3. \emph{Density-Ratio estimation}} % This is for the header on each page - perhaps a shortened title

%----------------------------------------------------------------------------------------
\section{Outline}
\begin{itemize}
  \item In-depth analysis on one of the methods of Chapter \ref{Chapter2}
  \item This method (e.g. Density-Ratio estimation) will be the basis for the real research
  \item Explain why this methods seems worthy and interesting
  \item Look at problems when applied to accelerometer sensor data
  \item \emph{The problems discovered here will give rise to the problem statement at the Introduction / beginning of research}
  \item Opens the possibility for own method
\end{itemize}

### 2013-05-08 Chapters/Chapter4.tex
New template, merged old text
Create outlines for new focus: temporal segmentation

% Chapter 3

\chapter{Experiments} % Main chapter title

\label{Chapter3} % For referencing the chapter elsewhere, use \ref{Chapter1}

\lhead{Chapter 3. \emph{Experiments}} % This is for the header on each 
%page - 
%perhaps a shortened title

%----------------------------------------------------------------------------------------

\section{}

>>>>

% !TEX root = ../main.tex
% Chapter 4

\chapter{Proposed method}

\label{Chapter4} % For referencing the chapter elsewhere, use \ref{Chapter4}

\lhead{Chapter 4. \emph{Proposed method}} % This is for the header on each page - perhaps a shortened title

%----------------------------------------------------------------------------------------
\section{Outline}
\begin{itemize}
  \item Based on the problem statement with current research as stated in Chapter \ref{Chapter3}
  \item Adjust method to needs
  \item Explain using graphs, pseudo-algorithms. Make clear distinction in origin of ideas and why to apply
\end{itemize}
